### 2024-08-20 parts/correction.tex
typo
test

First, let us remark that $\CorrectPoly \subseteq \CoveredPoly$, because
polynomials in $\CorrectPoly$ are \kl{non-negative}. We also remarked at the
beginning of this section that our counterexample provided in
\cref{thm:counter-example} is not in $\CorrectPoly$.

\AP Let us now prove that \kl{$\Nat$-rational polynomials} are in
$\CorrectPoly$. This follows from the correct implication in the statement of
\cref{karh:thm}, but we provide a self-contained proof using a refinement of
the classical combinatorial arguments for $\ZPoly$ \cite[Lemma 4.16]{CDTL23}
and $\NPoly$ \cite[Lemma 5.37]{DOUE23}. We take extra care to reprove in
\cref{n-poly-combinatorics:lem} a finer statement that has two main goals:
first, it formulates the traditional \emph{pumping argument} as a composition
property using \kl{(noncommutative output) polynomial functions}, allowing us
to use the same formalism in \cref{k-combinatorial:def} and
\cref{ultimately-polynomial:def} and highlight the role of \kl{commutative}
\kl{polyregular functions} in the broader study of \kl{polyregular functions};
then we also give a precise shape of the polynomials that arise from such
\emph{pumping arguments}, which was lacking in former similar statements where
$\Rat[X_1, \dots, X_n]$ was used as an over approximation.

\AP To address our first goal, let us define that a function $q$ is a
\intro{(noncommutative output) polynomial function} from $\Nat^p$ to $\Sigma^*$
whenever there exists words $\alpha_0, \dots, \alpha_p \in \Sigma^*$, and words
$u_1, \dots, u_p \in \Sigma^*$, such that $q(X_1, \dots, X_p) = \alpha_0
\prod_{i = 1}^p u_i^{X_i} \alpha_i$. That is, $q$ is syntactically defined by a
non-commutative monomial over the semi- $\Sigma^*$. Almost by
definition, \kl{polynomial functions} are \kl{commutative} \kl{polyregular
functions}.

@@

Let us highlight that the multiplicative factor $\omega$ is necessary in
\cref{n-poly-combinatorics:lem}. Indeed, the function $f \colon \set{a}^* \to
\Nat$ defined as $0$ when the input is of odd length and $1$ when the input is
of even length is \kl{$\Nat$-polyregular}, but $f(a^X)$ is not a polynomial.
We can trade-off this multiplicative factor for a constant term 
addition under the extra assumption that the function 
is \kl{star-free polyregular}, as described in the following lemma.

@@

Let us also address the fact that the distinction between \kl{natural binomial
functions} and \kl{natural binomial polynomials} vanishes when we are
interested in polynomials. As a consequence, to a \kl{natural binomial
function} corresponds exactly one \kl{natural binomial polynomial} that
coincides with it on sufficiently large inputs.

@@

\begin{theorem}
    \label{corrected-version:thm}
    Let $P \in \Rel[\vec{X}]$.
    The following are equivalent:
    \begin{enumerate}
        \item \label{corrected-1:item} $P \in \CorrectPoly$,
        \item \label{corrected-2:item} $P$ is \kl{represented} by a \kl{$\Nat$-rational series},
        \item \label{corrected-3:item} $P$ is \kl{represented} by a \kl{$\Nat$-polyregular function},
        \item \label{corrected-4:item} $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function},
    \end{enumerate}
    Furthermore, the properties are decidable, and conversions effective.
\end{theorem}
\begin{proof}
    The implications 
    \cref{corrected-4:item} $\implies$
    \cref{corrected-3:item} $\implies$
    \cref{corrected-2:item} are obvious.
    \cref{lem:correct-to-n-rat} proves
    \cref{corrected-1:item} $\implies$ \cref{corrected-4:item},
    while \cref{n-rat-correct:lem}
    proves 
    \cref{corrected-2:item} $\implies$ \cref{corrected-1:item}.
    Note that the lemmas provide effective conversion procedures,
    and that
    \cref{derivation-translation:lem}
    also provides a decision
    procedure.

@@

\begin{corollary}
    \label{decide-rat-poly-npoly:cor}
    Let $P \in \Rat[\vec{X}]$ be a polynomial with \emph{rational}
    coefficients and let $\alpha$ be the smallest number in $\Nat_{\geq 1}$
    such that $\alpha P \in \Rel[\vec{X}]$. Then, the following are equivalent:
    \begin{enumerate}
        \item \label{rat-npoly-0:item} $P$ is a \kl{non-negative} \kl{natural binomial polynomial},
        \item \label{rat-npoly-1:item} $P$ is a \kl{natural binomial polynomial} such that $P(0,\dots,0) \geq 0$,
        \item \label{rat-npoly-2:item} $P$ is a \kl{$\Nat$-rational polynomial},
        \item \label{rat-npoly-3:item} $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function},
        \item \label{rat-npoly-4:item} $\alpha P \in \CorrectPoly$,
            and for all $\vec{x} \in \Nat^{\vec{X}}$,
            $P(\vec{x})$ is a multiple of $\alpha$.
    \end{enumerate}
    In particular, the properties are decidable.
\end{corollary}
\begin{proof}

    Proof of \cref{rat-npoly-4:item} $\implies$ \cref{rat-npoly-3:item}.
    If $\alpha P \in \CorrectPoly$, then  it is an \kl{$\Nat$-rational polynomial}.
    But in particular, there exists a polyregular function $f$ that represents $\alpha P$
    with outputs in $\set{ a }$.
    Now, let us post-compose this function by the one that
    maps $a^{k \alpha + r}$ to $a^{k}$, which is a \kl{polyregular function}.
    We conclude that $P$ is \kl{represented} by an \kl{$\Nat$-polyregular function}.

    Proof of \cref{rat-npoly-1:item} $\implies$ \cref{rat-npoly-0:item}.
    If $P$ is a \kl{natural binomial polynomial} with $P(0,\dots,0) \geq 0$,
    then it is \kl{non-negative}.

    Proof of \cref{rat-npoly-0:item} $\implies$ \cref{rat-npoly-3:item}.
    If $P$ is a \kl{non-negative} \kl{natural binomial polynomial},
    then for large enough values, it is a \kl{natural binomial function},
    hence computable, and for small values we can hardcode.

    Proof of \cref{rat-npoly-2:item} $\implies$ \cref{rat-npoly-4:item}.
    If $P$ is a \kl{$\Nat$-rational polynomial}, then $\alpha P$ is one too,
    and in particular $\alpha P$ belongs to $\CorrectPoly$.

    Proof of \cref{rat-npoly-3:item} $\implies$ \cref{rat-npoly-2:item}.
    This is because $\NSF \subseteq \NPoly$.

    Proof of \cref{rat-npoly-3:item} $\implies$ \cref{rat-npoly-0:item}.
    Because $P$ is computed by a function $f \in \NSF$,
    one can leverage \cref{n-sf-combinatorics:lem}
    to conclude that $P$ is actually a \kl{natural binomial polynomial}.
\end{proof}

@@

\begin{corollary}
    \label{integer-binomial-polynomials:cor}
    Let $P \in \Rat[\vec{X}]$. Then, $P$ is a \kl{$\Rel$-rational polynomial}
    if and only if $P$ is an \kl{integer binomial polynomial}.
\end{corollary}
\begin{proof}
    Let $f = f_+ - f_-$ where both $f_+$ and $f_-$ are 
    \kl{$\Nat$-polyregular functions}.
    Let us argue that $f_+$ and $f_-$ are \kl{star-free polyregular functions},
    because they are \kl{ultimately polynomial}. (todo later).
    As a consequence, one can apply \cref{n-sf-combinatorics:lem}
    to conclude that $P$ coincides with an 
    \kl{integer binomial polynomial} over large enough inputs,
    hence that $P$ is such a polynomial.
\end{proof}

>>>>

First, let us remark that $\CorrectPoly \subseteq \CoveredPoly$, because
polynomials in $\CorrectPoly$ are \kl{non-negative}. We also remarked at the
beginning of this section that our counterexample $\BadPoly$ provided in
\cref{thm:counter-example} is not in $\CorrectPoly$. The rest of the
section is mainly concerned with proving the following corrected version of
\cref{karh:thm}.

\begin{theorem}[restate=corrected-version:thm,label=corrected-version:thm]
    Let $P \in \Rel[\vec{X}]$.
    The following are equivalent:
    \begin{enumerate}
        \item \label{corrected-1:item} $P \in \CorrectPoly$,
        \item \label{corrected-2:item} $P$ is \kl{represented} by a \kl{$\Nat$-rational series},
        \item \label{corrected-3:item} $P$ is \kl{represented} by a \kl{$\Nat$-polyregular function},
        \item \label{corrected-4:item} $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function},
    \end{enumerate}
    Furthermore, the properties are decidable, and conversions effective.
\end{theorem}

\AP Let us now prove that \kl{$\Nat$-rational polynomials} are in
$\CorrectPoly$. This follows from the correct implication in the statement of
\cref{karh:thm}, but we provide a self-contained proof using a refinement of
the classical combinatorial arguments for $\ZPoly$ \cite[Lemma 4.16]{CDTL23}
and $\NPoly$ \cite[Lemma 5.37]{DOUE23}. We take extra care to reprove in our
upcoming \cref{n-poly-combinatorics:lem} a finer statement that has two main
goals: first, it formulates the traditional \emph{pumping argument} as a
composition property using \kl{(noncommutative output) polynomial functions},
allowing us to use the same formalism in the upcoming
\cref{k-combinatorial:def,ultimately-polynomial:def} and highlight the role of
\kl{commutative} \kl{polyregular functions} in the broader study of
\kl{polyregular functions}; then we also give a precise shape of the
polynomials that arise from such \emph{pumping arguments}, which was lacking in
former similar statements where $\Rat[X_1, \dots, X_n]$ was used as an over
approximation.

\AP To address our first goal, let us define that a function $q$ is a
\intro{(noncommutative output) polynomial function} from $\Nat^p$ to $\Sigma^*$
whenever there exists words $\alpha_0, \dots, \alpha_p \in \Sigma^*$, and words
$u_1, \dots, u_p \in \Sigma^*$, such that $q(X_1, \dots, X_p) = \alpha_0
\prod_{i = 1}^p u_i^{X_i} \alpha_i$. That is, $q$ is syntactically defined by a
non-commutative monomial over the monoid $\Sigma^*$. Almost by
definition, \kl{polynomial functions} are \kl{commutative} \kl{polyregular
functions}.

@@

Let us highlight that the multiplicative factor $\omega$ is necessary in
\cref{n-poly-combinatorics:lem}. Indeed, the function $f \colon \set{a}^* \to
\Nat$ defined as $0$ when the input is of odd length and $1$ when the input is
of even length is \kl{$\Nat$-polyregular}, but $f(a^X)$ is not a polynomial. We
can trade off this multiplicative factor for a constant term addition under the
extra assumption that the function is \kl{star-free polyregular}, as described
in the following \cref{n-sf-combinatorics:lem}.

@@

Let us also address the fact that the distinction between \kl{natural binomial
functions} and \kl{natural binomial polynomials} vanishes when we are
interested in polynomials. As a consequence, to a \kl{natural binomial
function} corresponds exactly one \kl{natural binomial polynomial} that
coincides with it on sufficiently large inputs (\cref{natural-binomial-polynomial:lem})
which allows us to conclude that \kl{$\Nat$-rational polynomials} are in
$\CorrectPoly$ (\cref{n-rat-correct:lem}).

@@

\begin{proofof}[corrected-version:thm]
    The implications 
    \cref{corrected-4:item} $\implies$
    \cref{corrected-3:item} $\implies$
    \cref{corrected-2:item} are obvious.
    \cref{lem:correct-to-n-rat} proves
    \cref{corrected-1:item} $\implies$ \cref{corrected-4:item},
    while \cref{n-rat-correct:lem}
    proves 
    \cref{corrected-2:item} $\implies$ \cref{corrected-1:item}.
    Note that the lemmas provide effective conversion procedures,
    and that
    \cref{derivation-translation:lem}
    also provides a decision
    procedure.
\end{proofof}

@@

\begin{lemma}
    \label{natural-binomia-correct:lem}
    Let $P$ be a \kl{non-negative} \kl{natural binomial polynomial},
    and let $\alpha$ be such that $\alpha P \in \Rel[\vec{X}]$.
    Then, $\alpha P \in \CorrectPoly$.
\end{lemma}
\begin{proof}
    The only think to remark is that
    $\restr{P}{\nu}$ remains a \kl{natural binomial polynomial}
    for all partial valuations $\nu$, and furthermore remains positive.
\end{proof}

\begin{corollary}
    \label{decide-rat-poly-npoly:cor}
    Let $P \in \Rat[\vec{X}]$ be a polynomial with \emph{rational}
    coefficients and let $\alpha$ be the smallest number in $\Nat_{\geq 1}$
    such that $\alpha P \in \Rel[\vec{X}]$. Then, the following are equivalent:
    \begin{enumerate}
        \item \label{rat-npoly-0:item} $P$ is a \kl{natural binomial polynomial} such that $P(0,\dots,0) \geq 0$,
        \item \label{rat-npoly-1:item} $P$ is a \kl{non-negative} \kl{natural binomial polynomial},
        \item \label{rat-npoly-2:item} $P$ is a \kl{$\Nat$-rational polynomial},
        \item \label{rat-npoly-3:item} $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function},
        \item \label{rat-npoly-4:item} $\alpha P \in \CorrectPoly$,
            and for all $\vec{x} \in \Nat^{\vec{X}}$,
            $P(\vec{x})$ is a multiple of $\alpha$.
    \end{enumerate}
    In particular, the properties are decidable.
\end{corollary}
\begin{proof}
    First, let us remark that 
    \cref{rat-npoly-0:item} $\iff$ \cref{rat-npoly-1:item}
    because \kl{natural binomial polynomials} are increasing.
    Furthermore, \cref{rat-npoly-3:item} $\implies$ \cref{rat-npoly-2:item}
    because $\NSF \subseteq \NPoly$.

    Then, \cref{rat-npoly-4:item} $\implies$ \cref{rat-npoly-3:item}, because
    if $\alpha P \in \CorrectPoly$, then it is an \kl{$\Nat$-rational
    polynomial}. But in particular, there exists a polyregular function $f$
    that represents $\alpha P$ with outputs in $\set{ a }$. Now, let us
    post-compose this function by the one that maps $a^{k \alpha + r}$ to
    $a^{k}$, which is a \kl{polyregular function}. We conclude that $P$ is
    \kl{represented} by an \kl{$\Nat$-polyregular function}.

    We also know that \cref{rat-npoly-2:item} $\implies$
    \cref{rat-npoly-4:item}, since if $P$ is an \kl{$\Nat$-rational
    polynomial}, then so is $\alpha P$, and in particular $\alpha P \in
    \CorrectPoly$ by \cref{corrected-version:thm}.

    As a consequence, we have proven that \cref{rat-npoly-0:item} $\iff$
    \cref{rat-npoly-1:item} and \cref{rat-npoly-2:item} $\iff$
    \cref{rat-npoly-3:item} $\iff$ \cref{rat-npoly-4:item}.

    Finally, \cref{rat-npoly-1:item} $\implies$ \cref{rat-npoly-4:item} is
    exactly \cref{natural-binomia-correct:lem}. And \cref{rat-npoly-3:item}
    $\implies$ \cref{rat-npoly-1:item} is obtained
    by combining \cref{n-sf-combinatorics:lem} and \cref{natural-binomial-polynomial:lem}.
\end{proof}

@@

\begin{corollary}
    \label{integer-binomial-polynomials:cor}
    Let $P \in \Rat[\vec{X}]$. Then, $P$ is a \kl{$\Rel$-rational polynomial}
    if and only if $P$ is an \kl{integer binomial polynomial}.
\end{corollary}
\begin{proof}
    Let $f = f_+ - f_-$ where both $f_+$ and $f_-$ are 
    \kl{star-free $\Nat$-polyregular functions}.
    One can apply \cref{n-sf-combinatorics:lem}
    to conclude that $P$ coincides with an 
    \kl{integer binomial polynomial} over large enough inputs,
    hence that $P$ is such a polynomial.
\end{proof}

### 2024-08-20 parts/intro.tex
test

\subparagraph*{Outline of the paper.} In \cref{preliminaries:sec}, we introduce
\kl{$\Nat$-polyregular functions} (resp. $\Rel$-polyregular functions), show
that one can decide if a function $f \in \ZPoly$ is \kl{commutative}
(\cref{decidable-commutative-poly:lemma}), and give the basic definitions used
on multivariate polynomials. In \cref{polynomials:sec}, we focus on
\emph{polynomials}, by introducing the \cref{karh:thm} of \cite{KARH77},
providing a counter-example in \cref{thm:counter-example}, and an alternative
characterization in \cref{corrected-version:thm}. Also in
\cref{polynomials:sec}, we provide characterizations of $\Rat$-polynomials that
can be expressed respectively by \kl{$\Nat$-polyregular functions}
(\cref{corrected-version:thm} \cref{corrected-0:item}) and
\kl{$\Rel$-polyregular functions} (\cref{integer-binomial-polynomials:cor}).
Finally, in \cref{beyond-polynomials:sec}, we leverage the results over
polynomials to decide $\NPoly$ inside $\ZPoly$ in the \kl{commutative} case
(\cref{decidable-n-poly:thm}), and prove that \cref{zsf-nsf:conjecture}
\cite[Conjecture 7.61]{DOUE23} holds for \kl{commutative} functions
(\cref{zsf-npoly-nsf:thm}).

>>>>

\subparagraph*{Outline of the paper.} In \cref{preliminaries:sec}, we provide
a combinatorial definition of
\kl{$\Nat$-polyregular functions} (resp. $\Rel$-polyregular functions), show
that one can decide if a function $f \in \ZPoly$ is \kl{commutative}
(\cref{decidable-commutative-poly:lemma}), and give the basic definitions used
on multivariate polynomials. In \cref{polynomials:sec}, we focus on
\emph{polynomials}, by introducing the \cref{karh:thm} of \cite{KARH77},
providing a counter-example in \cref{thm:counter-example}, and an alternative
characterization in \cref{corrected-version:thm}. Also in
\cref{polynomials:sec}, we provide characterizations of $\Rat$-polynomials that
can be expressed respectively by \kl{$\Nat$-polyregular functions}
(\cref{decide-rat-poly-npoly:cor}) and
\kl{$\Rel$-polyregular functions} (\cref{integer-binomial-polynomials:cor}).
Finally, in \cref{beyond-polynomials:sec}, we leverage the results over
polynomials to decide $\NPoly$ inside $\ZPoly$ in the \kl{commutative} case
(\cref{decidable-n-poly:thm}), and prove that \cref{zsf-nsf:conjecture}
\cite[Conjecture 7.61]{DOUE23} holds for \kl{commutative} functions
(\cref{zsf-npoly-nsf:thm}).

### 2024-08-20 parts/starfree.tex
test

\AP The decidability of aperiodicity for \kl{$\Rel$-polyregular functions}, that is
the membership of $\ZSF$ inside $\ZPoly$,
relies on the construction of a canonical object called the \emph{residual
,
transducer}, the latter being essentially based on differences between
functions, crucially leveraging negative outputs \cite{CDTL23}. Although the
proof method does not carry from $\Rel$-output functions to $\Nat$-output,
functions, it was conjectured that the semantic property of being
\kl{ultimately polynomial} would also characterize $\NSF$ inside $\NPoly$, a
conjecture that is restated hereafter.

>>>>

\AP The decidability of aperiodicity for \kl{$\Rel$-polyregular functions}, that is
the membership of $\ZSF$ inside $\ZPoly$,
relies on the construction of a canonical object called the \emph{residual
transducer}, the latter being essentially based on differences between
functions, crucially leveraging negative outputs \cite{CDTL23}. Although the
proof method does not carry from $\Rel$-output functions to $\Nat$-output,
functions, it was conjectured that the semantic property of being
\kl{ultimately polynomial} would also characterize $\NSF$ inside $\NPoly$, a
conjecture that is restated hereafter.
